Analysing Constraint Grammar with a SAT solver

### 2015-10-13 paper/lrec/implementation.tex
more text

\section{Implementation}

In this section, we describe the implementation of the toolkit.

@@

\begin{table*}[]
\centering
\caption{My caption}
\label{my-label}
\begin{tabular}{|l|p{1.9cm}|l|l|}
\hline
\textbf{Word} & \textbf{Variables} & \textbf{Default rules} & \textbf{Constraint rules}\\ \hline
\multirow{2}{*} ``the''  & \texttt{the\_det}  & \texttt{the\_det} & {\texttt{the\_det} $\Rightarrow$ $\neg$\texttt{bear\_v\_pl}} \\
                ``bear'' & \texttt{bear\_n\_sg}  \texttt{bear\_v\_pl} & \texttt{bear\_n\_sg} $\vee$ \texttt{bear\_v\_pl} & \\  \hline
\end{tabular}

@@

Given a large morphological lexicon, it is relatively easy to restrict each individual word to contain only realistic ambiguities. We expand the lexicon, map each wordform to its possible analyses, and then ignore the actual wordforms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:<verb><sg><p3>} and \texttt{casa:<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.
However, since our method is corpus-free, it does not contain any restrictions as to what words can follow each other. \footnote{Encode morphological lexicon and CG rules in SAT + ask for a sentence that will trigger rules = natural language generation! :D}

>>>>

\section{Implementation}
\label{sec:implementation}

In this section, we describe the implementation of the toolkit.
We encode Constraint Grammar as a SAT problem, as in \cite{listenmaa_claessen2015}, and from the morphological lexicon, we get constraints for individual words. This allows us to create word sequences that will or will not trigger 
use a SAT solver to constrain the possibilities.

@@

\begin{table*}[]
\centering
\caption{SAT-encoding of CG}
\label{my-label}
\begin{tabular}{|l|p{1.9cm}|l|l|}
\hline
\textbf{Word} & \textbf{Variables} & \textbf{Default rules} & \textbf{Constraint rules}\\ \hline
\multirow{2}{*} ``the''  & \texttt{the\_det}  & \texttt{the\_det} & {\texttt{the\_det} $\Rightarrow$ $\neg$\texttt{bear\_v\_pl}} \\
                ``bear'' & \texttt{bear\_n\_sg}  \texttt{bear\_v\_pl} & \texttt{bear\_n\_sg} $\vee$ \texttt{bear\_v\_pl} & \\  \hline
\end{tabular}

@@

Given a large morphological lexicon, it is relatively easy to restrict each individual word to contain only realistic ambiguities. We expand the lexicon, map each wordform to its possible analyses, and then ignore the actual wordforms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:<verb><sg><p3>} and \texttt{casa:<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.
However, since our method is corpus-free, it does not contain any restrictions as to what words can follow each other.
% \footnote{Encode morphological lexicon and CG rules in SAT + ask for a sentence that will trigger rules = natural language generation! :D}

### 2015-10-13 paper/lrec/problem_desc.tex
more text

As CGs grow larger, .

\cite{bick2013tuning} presents efforts to optimise hand-written CGs using machine learning techniques.
While Bick's experiments show improvements in results, the grammar writer is none the wiser why is the grammar better.
Our toolset is designed to complement the existing efforts.
The ideal use case would be a grammar writer, who wants to know e.g.

\begin{itemize}
\item Is there a sentence that triggers rule(s) X but not rule(s) Y?
\item Give a sentence that is ambiguous but doesn't trigger any rules
\item Does my grammar contain rules that contradict each other?
\item Does my grammar contain rules that will never fire?
\end{itemize}

The technique requires an existing morphological lexicon, but no corpus.
This makes it suitable also for languages without large corpora.
Authors writing grammars for well-resourced languages can also benefit from the technique, as an additional tool along with a gold standard corpus. 
Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin (\cite{voutilainen2004}). For a language with no free hand-tagged corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to Wikipedia texts and pick 100 examples at random for manual check.

\section{Previous work}

This work draws inspiration from various sources.

\begin{itemize}
\item Methods for grammar writing \cite{voutilainen2004},  \cite{tyers_reynolds2015}
\item Optimising a hand-written CG \cite{bick2013tuning}
\item Encoding CG as a logic program \cite{lager98}, \cite{lindberg_eineborg98ilp},  \cite{lager01transformation}, \cite{asfrent14}, \cite{listenmaa_claessen2015}
\end{itemize}

Things you can get from corpus:

\begin{itemize}
\item \# times the rule was applied
\item \% times it was correct
\end{itemize}
Strengths: gives concrete feedback about the effect of rules. 
Good especially at the beginning stages of grammar writing. 
Sometimes you just want to know that a rule fired 500 times and was correct 490 times, rather than ``it is possible to construct a word sequence that will trigger this rule''.
The connection to the reality has also weak sides. This method will not notice a missed opportunity. What if instead of firing 500 times, it could have fired 1000 times if it had been placed earlier in the rule sequence.
Or the rule that was wrong in 60 \% of the time,

% \cite{voutilainen2004} state that the first 200 rules are probably enough to resolve 50--75 \% of ambiguities in the corpus used in the development.

Things you can get from ML tuning:
\begin{itemize}
\item Variations in the place of the rule
\item Variations in the conditions (careful, ...) of the rule
\end{itemize}
Good side: try all combinations to find the best result. 
At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.

Things you can get from SAT solver:
\begin{itemize}
\item Interaction of rules
\item Create word sequences to test what is possible and what can never happen
\end{itemize}

>>>>

As CGs grow larger, it gets harder for the grammar writers and users to keep track of all the rules and their interaction.
\cite{bick2013tuning} presents efforts to optimise hand-written CGs using machine learning techniques.
While Bick's experiments show improvements in results, the grammar writer is none the wiser why is the grammar better.
Our toolset is designed to complement the existing approaches in CG analysis and optimisation.
The ideal use case would be a grammar writer, who wants to know e.g.

\begin{itemize}
\item Does my grammar contain rules that contradict each other?
\item Does my grammar contain rules that will never fire?
\item Generate a sequence that triggers rule(s) X but not rule(s) Y
\item Generate a sequence that is ambiguous but doesn't trigger any rules
\end{itemize}

The technique requires an existing morphological lexicon, but no corpus. 
The lexicon is needed in order to generate sequences that have realistic ambiguities: e.g. a word can be ambiguous between noun and verb, but not between adjective and punctuation. 
Constraints for the structure of the sequences come from the CG rules themselves.
We encode both of these constraints as a satisfiability problem, and use a SAT solver.

The paper is structured as follows. Section~\ref{sec:prev} relates our work to the previous research. Section~\ref{sec:implementation} discusses the techniques we use, Section~X presents preliminary results and  Section~Y discusses future work. Section~Z concludes the paper.

\section{Previous work}
\label{sec:prev}

Our work draws inspiration from various sources. This section explains what our approach is and is not, and relates it to the previous research. We combine elements from the following aspects of CG research:

\begin{itemize}
\item Methods to aid in hand-crafting rules \cite{voutilainen2004}, \cite{tyers_reynolds2015}
\item Optimising a hand-written CG \cite{bick2013tuning}
\item Encoding CG in logic \cite{lager98}, \cite{lager_nivre01}, \cite{lager01transformation}, \cite{listenmaa_claessen2015}
\end{itemize}

There is a large body of research on automatically inducing rules, e.g. \cite{inducing_cg1996}, \cite{lindberg_eineborg98ilp}%, \cite{asfrent14}
. Our work is aimed to aid the process of hand-crafting rules.

\paragraph{Corpus-based methods in manual grammar development}

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}. For a language with no free hand-tagged corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to Wikipedia texts and pick 100 examples at random for manual check.
\begin{itemize}
\item \# times the rule was applied
\item \% times it was correct
\end{itemize}
Strengths: gives concrete feedback about the effect of rules. 
Good especially at the beginning stages of grammar writing. 
Sometimes you just want to know that a rule fired 500 times and was correct 490 times, rather than ``it is possible to construct a word sequence that will trigger this rule''.
The connection to the reality has also weak sides. This method will not notice a missed opportunity. What if instead of firing 500 times, it could have fired 1000 times if it had been placed earlier in the rule sequence.
Or the rule that was wrong in 60 \% of the time,

% \cite{voutilainen2004} state that the first 200 rules are probably enough to resolve 50--75 \% of ambiguities in the corpus used in the development. 
Authors writing grammars for well-resourced languages can also benefit from the technique, as an additional tool along with a gold standard corpus.

\paragraph{ML optimisation of hand-written grammars}
\begin{itemize}
\item Variations in the place of the rule
\item Variations in the conditions (careful, ...) of the rule
\end{itemize}
Good side: try all combinations to find the best result. 
At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.

\paragraph{Things you can get from SAT solver}

\begin{itemize}
\item Interaction of rules
\item Create word sequences to test what is possible and what can never happen
\end{itemize}
